Competing Bandits: \\ The Perils of Exploration under Competition

### 2019-10-22 journal paper/content/related_work.tex
updates

A line of work on \emph{platform competition}, starting with \cite{Rysman09}, concerns competition between firms (\emph{platforms}) that improve as they attract more users (\emph{network effect}); see \citet{Weyl-White-14} for a recent survey. This literature is not concerned with \innovation, and typically models network effects exogenously, whereas in our model network effects are endogenous: they are created by MAB algorithms, an essential part of the model. \gaedit{The ``death spiral" effect that we find is similar to the increasing dominance dominance notion found in the industrial organization literature (see e.g. \cite{cabral1994learning, athey2001investment}). Further, there is also a line of work that investigates how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers \cite{schmalensee1982product, bagwell1990informational}. In the reputation choice model, we observe a similar effect when we investigate the role that reputation can serve as a barrier to entry. However, in our model this effect is further strengthened by the fact that the firms have to learn while competing adding that the incumbent may not only have a reputational advantage but additionally the information it acquires serves as a further barrier to entry. The idea that data can serve as a barrier to entry, especially in online markets, has been studied theoretically in \cite{ lambrecht2015can} and empirically in
\cite{bajari2018impact}. While these papers find that small amounts of additional data do not provide significant improvement, they focus on learning in isolation. }

>>>>

A line of work on \emph{platform competition}, starting with \cite{Rysman09}, concerns competition between firms (\emph{platforms}) that improve as they attract more users (\emph{network effect}); see \citet{Weyl-White-14} for a recent survey. This literature is not concerned with \innovation, and typically models network effects exogenously, whereas in our model network effects are endogenous: they are created by MAB algorithms, an essential part of the model. \gaedit{The ``death spiral" effect that we find is similar to the increasing dominance notion found in the industrial organization literature (see e.g. \cite{cabral1994learning, athey2001investment}). Further, there is also a line of work that investigates how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers \cite{schmalensee1982product, bagwell1990informational}. In the reputation choice model, we observe a similar effect when we investigate the role that reputation can serve as a barrier to entry. However, in our model this effect is further strengthened by the fact that the firms have to learn while competing adding that the incumbent may not only have a reputational advantage but additionally the information it acquires serves as a further barrier to entry. The idea that data can serve as a barrier to entry, especially in online markets, has been studied theoretically in \cite{ lambrecht2015can} and empirically in
\cite{bajari2018impact}. While these papers find that small amounts of additional data do not provide significant improvement, they focus on learning in isolation. }

### 2019-10-22 journal paper/content/sec-intro.tex
updates

Learning from interactions with users is ubiquitous in modern customer-facing systems, from product recommendations to web search to spam detection to content selection to fine-tuning user interfaces. Many systems purposefully implement \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. Randomized controlled trials, a.k.a. A/B testing, are an industry standard, with a number of companies such as \emph{Optimizely} offering tools and platforms to facilitate them. Many companies use more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known theoretical framework for exploration and making decisions under uncertainty.

%~\cite{KohaviAB-2015,KohaviLSH09}

Systems that engage in exploration typically need to compete against one another; most importantly, they compete for users. This creates an interesting tension between \exploration and \competition. In a nutshell, while exploring may be essential for improving the service tomorrow, it may degrade quality and make users leave \emph{today}, in which case there will be no users to learn from! Thus, users play three distinct roles: they are customers that generate revenue, they generate data for the systems to learn from, and they are self-interested agents which choose among the competing systems.

We initiate a study of the interplay between \exploration and \competition. The main high-level question is: {\bf whether and to what extent competition incentivizes adoption of better exploration algorithms}. \gaedit{This translates into a number of more concrete questions. While it is commonly assumed that better learning technology always helps, is this so for our setting? Does increasing competition lead to higher consumer welfare? To what extent does one firm possessing more information or having a better perception amongst consumers than its competitors serve as a barrier to entry in this environment?

We investigate these questions for two separate but connected stylized duopoly models where two firms commit to exploration strategies and compete for a stream of consumers. In the first, which we denote as the \textit{expectation choice model}, we suppose that consumers are well-informed about the firm's exploration strategies but do not observe any information about past user experiences with the firms. In the second, which we denote as the \textit{reputation choice model}, we suppose that consumers \textit{only} consider the average of past user experiences with each of the firms. In the expectation choice model we analytically characterize the equilibrium exploration strategies as we vary the degree of randomness in consumer decisions. In the reputation choice model we account for competition in a more direct way and utilize numerical simulations to consider asymmetries in the timing of entry, increasing competition beyond two firms, and allow us to isolate the effect that role that informational advantages can play as a barrier to entry in markets where exploration and competition tensions arise.

@@

\item Principals simultaneously announce their learning algorithms before round $1$, and cannot change them afterwards. There is a common Bayesian prior on the rewards (but the realized reward distributions are not observed by the principals or the agents).  Each principal only observes agents that chose him. \gaedit{The information set for the agent differs between the two variants of the model. \begin{enumerate}
\item In the expectation choice model, agents do not receive any other information and choose between the principals using their knowledge of $t$ and the principals' algorithms
\item In the reputation choice model, agents have access to a reputation score for each principal, which is a sliding window average of the rewards experienced by previous agents that have visited this principal
\end{enumerate}
}
\end{itemize}

\gaedit{
\xhdr{Main Findings.}
In both variants of our model, we find that in a simultaneous entry duopoly a greedy algorithm that does no purposeful exploration is incentivized in equilibrium. In the expectation choice model, once a firm does any exploration it is immediately starved of consumers when its opponent plays a greedy algorithm since we show that the expected reward for every subsequent consumer is higher for the firm that plays the greedy algorithm. In the reputation choice model, the same mechanism drives the greedy algorithm to be incentivized in equilibrium. In this variant, exploration hurts a firm's reputation and decrease its market share in the near term. This leaves the firm with less users to learn from, which further degrades the firm's performance relative to competitors who keep learning and improving from \emph{their} customers, and so forth. Taken to the extreme, such dynamics lead to a ``death spiral" effect when the vast majority of customers eventually switch to competitors. In both cases, consumer welfare is low since the greedy algorithm is known to be dramatically bad in many important cases of multi-armed bandits.

The primary mechanism that generates this stark result is the same in both variants of the model: consumers need to be incentivized to select a firm over its competitors, leading firms that engage in exploration to be starved of consumers before they make enough progress on their learning problem. In order to incentivize ``better" exploration strategies the key intuition is that the firm needs to have some ``free" consumers that visit them without the firm having to incentivize them to do so. This allows the firm to eventually overcome the initial losses in consumer perception from exploration.

In the expectation choice model, we relax the decision rule of the consumers to allow for each firm to be chosen with some fixed baseline probability so that firms get a constant stream of ``free" consumers; we call this choice rule \HardMaxRandom. We find that, in this setting, better algorithms help in a big way: a sufficiently better algorithm is guaranteed to win all non-random agents after an initial learning phase. While the precise notion of ``sufficiently better algorithm" is rather subtle, we note that commonly known ``smart" bandit algorithms typically defeat the commonly known ``naive" ones, and the latter typically defeat the greedy algorithm. However, there is a substantial caveat: one can defeat any algorithm by interleaving it with the greedy algorithm. This has two undesirable corollaries: a better algorithm may sometimes lose, and a pure Nash equilibrium typically does not exist.

In the reputation choice model, we maintain the same decision rule for the consumers but instead vary the timing of entry and allow one firm to have a first-mover advantage so that this firm (denoted as the incumbent) gets all the agents in the market before the other firm enters. We find that, if the first-mover period is sufficiently large, then the incumbent has a dominant strategy to play ``smart" bandit algorithms and that consumer welfare is substantially higher than in the case of simultaneous entry. The intuition is simply that a sufficiently long incumbency period allows the firm to make sufficient progress on its learning problem that it can overcome the original drop in its reputation from exploration in the beginning rounds.

\xhdr{Additional Findings} In the expectation choice model, we further relax the decision rule of the agents so that the probability of choosing a given principal varies smoothly as a function of the difference between  principals' expected rewards; we call it \SoftMaxRandom. For this model, the ``better algorithm wins" result holds under much weaker assumptions on what constitutes a better algorithm. This is the most technical result of the paper. The competition in this setting is necessarily much more relaxed: typically, both principals attract approximately half of the agents as time goes by (but a better algorithm may attract slightly more).

In the reputation choice model, we investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate and compare these two effects. We find that either effect alone leads to a significant advantage under competition. The data advantage is larger than reputation advantage when the incumbent commits to a more advanced bandit algorithm.

@@

We also investigate how algorithms' performance ``in isolation" (without competition) is predictive of the outcomes under competition. We find that mean reputation -- arguably, the most natural performance measure ``in isolation" -- is sometimes not a good predictor. We suggest a
more refined performance measure, and use it to explain some of the competition outcomes.

@@

In our model, competition varies based on the number of ``free" consumers that a firm receives. In the expectation choice model, this comes from introducing ``random" consumers where we vary from fully rational decisions with \HardMax to relaxed rationality with \HardMaxRandom to an even more relaxed rationality with \SoftMaxRandom. Indeed, with \HardMax you lose all customers as soon as you fall behind in performance, with \HardMaxRandom you get some small market share no matter what, and with \SoftMaxRandom you are further guaranteed a market share close to $\tfrac12$ as long as your performance is not much worse than the competition. The uniform choice among principals corresponds to no rationality and no competition. In the reputation choice model, this comes from the number of firms and the timing of entry in the market. Competition in this case ranges from monopoly to ``incumbent" (first-mover in duopoly) to simultaneous entry duopoly to ``late entrant" (last mover in duopoly). The same distinctions in both cases also control the severity of competition between the principals.

@@

\xhdr{Map of the paper.}
We survey related work (Section~\ref{sec:related-work}), lay out the model and preliminaries (Section~\ref{sec:model}), and proceed to analyze the different models. We start by analyzing the expectation choice model analytically. In Sections ~\ref{sec:rational},~\ref{sec:random}, and ~\ref{sec:soft} we analyze the first main variant of the model and characterize the equilibrium behavior under three different agent decision rules. Then, we turn to analyze the reputation choice variant of the model with details of the analysis described in Section ~\ref{sec:sim_details}. Sections ~\ref{sec:isolation} and ~\ref{sec:revisited} overview results from running different bandit algorithms in isolation (i.e. without competition). Sections ~\ref{sec:competition} and ~\ref{sec:barriers} overview the results of the second variant of the mode. Section ~\ref{sec:conclusion} concludes.
}

>>>>

Learning from interactions with consumers is ubiquitous in modern customer-facing systems, from product recommendations to web search to content selection to fine-tuning user interfaces. These interactions are essential for the development of high-quality products, but, in order to learn what products are high quality, companies need to implement systems that engage in \emph{exploration}: deploying potentially suboptimal products in order to acquire new information to make more informed product decisions tomorrow.

In this paper, we initiate a study of the interplay between \exploration and \competition. While exploration may be beneficial for both firms and consumers in the long-run, it may lead to bad experiences for consumers today and result in subsequent consumers going to the firm's competitors tomorrow. In particular, a firm can only make progress on its learning problem if it is able to incentivize consumers to visit it over its competitors and excessive exploration may lead to low reputation which leads consumers to increasingly visit the firm's competitors. Consumers therefore serve three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents who choose among the competing firms.

Given the prevalence of these technologies and the increased importance of data in the digital economy, our main high-level question is: {\bf does competition lead to the adoption of better exploration algorithms?}. This translates into a number of more concrete questions. While it is commonly assumed that ``better" technology always helps, is this so for our setting? Does increasing competition lead to higher consumer welfare? To what extent is there a ``data feedback loop" where one firm having more data leads to that firm attracting more users which leads to that firm having more data, etc.?

We investigate these questions with a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We first look at a version of this model that allows us to make analytical characterizations of the optimal algorithms firms should deploy in equilibrium. In this variant of the model, which we denote as the \textit{expectation choice variant}, we suppose that consumers are well-informed about the firm's exploration strategies but do not observe any information about past user experiences with the firms. In this variant we analytically characterize the equilibrium exploration strategies as we vary the degree of randomness in consumer decisions. We then look at a second variant of the model that accounts for competition in a more direct way, which we denote as the \textit{reputation choice variant}, where we suppose that consumers \textit{only} consider the average of past user experiences with each of the firms. This variant allows us to utilize numerical simulations to consider a number of additional issues such as asymmetries in the timing of entry, increasing competition beyond two firms, and allowing us to isolate the role that informational advantages can play as a barrier to entry in markets where exploration and competition tensions arise.

@@

\item Principals simultaneously announce their learning algorithms before round $1$, and cannot change them afterwards. There is a common Bayesian prior on the rewards (but the realized reward distributions are not observed by the principals or the agents).  Each principal only observes agents that chose him. The information set for the agent differs between the two variants of the model. \begin{enumerate}
\item In the expectation choice variant, agents do not receive any other information and choose between the principals using their knowledge of $t$ and the principals' algorithms
\item In the reputation choice variant, agents have access to a reputation score for each principal, which is a sliding window average of the rewards experienced by previous agents that have visited this principal
\end{enumerate}
\end{itemize}

\xhdr{Main Findings.}
We find that in a simultaneous entry duopoly a greedy (myopic) algorithm that does no purposeful exploration is incentivized in equilibrium. In the expectation choice variant, once a firm does any exploration it is immediately starved of consumers when its opponent plays a greedy algorithm since we show that the expected reward for every subsequent consumer is higher for the firm that plays the greedy algorithm. In the reputation choice variant, the same mechanism drives the greedy algorithm to be incentivized in equilibrium. In this variant, exploration hurts a firm's reputation and decrease its market share in the near term. This leaves the firm with less users to learn from, which further degrades the firm's performance relative to competitors who keep learning and improving from \emph{their} customers, and so forth. Taken to the extreme, such dynamics lead to a ``death spiral" effect when the vast majority of customers eventually switch to competitors. As a result, our model provides an example of the mechanisms that can lead to a ``data feedback loop" that has been hypothesized in policy discussions. In both cases, consumer welfare is low since the greedy algorithm is known to be dramatically bad in many important cases of multi-armed bandits.

The primary mechanism that generates this stark result is that consumers need to be incentivized to select a firm over its competitors, leading firms that engage in exploration to be starved of consumers before they make enough progress on their learning problem. In order to incentivize ``better" exploration strategies the key intuition is that the firm needs to have some ``free" consumers that visit them without the firm having to incentivize them to do so. This allows the firm to eventually overcome the initial losses in consumer perception from exploration.

We relax the decision rule of the consumers to allow for each firm to be chosen with some fixed baseline probability so that firms get a constant stream of ``free" consumers; we call this choice rule \HardMaxRandom. We find that, in this setting, better algorithms help in a big way: a sufficiently better algorithm is guaranteed to win all non-random agents after an initial learning phase. While the precise notion of ``sufficiently better algorithm" is rather subtle, we note that commonly known ``smart" bandit algorithms typically defeat the commonly known ``naive" ones, and the latter typically defeat the greedy algorithm. However, there is a substantial caveat: one can defeat any algorithm by interleaving it with the greedy algorithm. This has two undesirable corollaries: a better algorithm may sometimes lose, and a pure Nash equilibrium typically does not exist.

We use the reputation choice variant to analyze the effect of varying the timing of entry between the firms and allow one firm to have a first-mover advantage so that this firm (denoted as the incumbent) gets all the agents in the market before the other firm enters. We find that, if the first-mover period is sufficiently large, then the incumbent has a dominant strategy to play ``smart" bandit algorithms and that consumer welfare is substantially higher than in the case of simultaneous entry. The intuition is simply that a sufficiently long incumbency period allows the firm to make sufficient progress on its learning problem that it can overcome the original drop in its reputation from exploration in the beginning rounds.

\xhdr{\underline{Additional Findings}}

\xhdr{Noise in consumer choice}: We further relax the decision rule of the agents so that the probability of choosing a given principal varies smoothly as a function of the difference between  principals' expected rewards; we call it \SoftMaxRandom. For this decision rule, the ``better algorithm wins" result holds under much weaker assumptions on what constitutes a better algorithm. This is the most technical result of the paper. The competition in this setting is necessarily much more relaxed: typically, both principals attract approximately half of the agents as time goes by (but a better algorithm may attract slightly more).

We use the reputation choice variant to explore numerically the extent to which better algorithms under the \HardMaxRandom decision rule are incentivized in realistic time-scales. We show that for small but still ``relevant" parameter values, this effect does not show up for unless the time horizon is very long. \\
\xhdr{Reputation and Data First-Mover Advantage}
In the reputation choice variant, we investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate and compare these two effects. We find that either effect alone leads to a significant advantage under competition. The data advantage is larger than reputation advantage when the incumbent commits to a more advanced bandit algorithm.

@@

\xhdr{Predicting Outcomes in Competition} We also investigate how algorithms' performance ``in isolation" (without competition) is predictive of the outcomes under competition in the reputation choice variant. We find that mean reputation -- arguably, the most natural performance measure ``in isolation" -- is sometimes not a good predictor. We suggest a more refined performance measure, based on a comparison between the reputation of the two firms, and use it to explain some of the competition outcomes.
\textbf{Add some discussion about comparison with ``better" algorithms in analytical part}

@@

In our model, competition varies based on the number of ``free" consumers that a firm receives. One economic mechanism that can generate this comes from introducing ``random" consumers where we vary from fully rational decisions with \HardMax to relaxed rationality with \HardMaxRandom to an even more relaxed rationality with \SoftMaxRandom. Indeed, with \HardMax you lose all customers as soon as you fall behind in performance, with \HardMaxRandom you get some small market share no matter what, and with \SoftMaxRandom you are further guaranteed a market share close to $\tfrac12$ as long as your performance is not much worse than the competition. The uniform choice among principals corresponds to no rationality and no competition. Another economic mechanism that can generate this comes from the number of firms and the timing of entry in the market. Competition in this case ranges from monopoly to ``incumbent" (first-mover in duopoly) to simultaneous entry duopoly to ``late entrant" (last mover in duopoly). The same distinctions in both cases also control the severity of competition between the principals.

@@

\xhdr{Map of the paper.}
We survey related work (Section~\ref{sec:related-work}), lay out the model and preliminaries (Section~\ref{sec:model}), and proceed to analyze the different models. We start by analyzing the expectation choice model analytically. In Sections ~\ref{sec:rational},~\ref{sec:random}, and ~\ref{sec:soft} we analyze the expectation choice variant and characterize the equilibrium behavior under three different agent decision rules. Then, we turn to analyze the reputation choice variant of the model with details of the analysis described in Section ~\ref{sec:sim_details}. Sections ~\ref{sec:isolation} and ~\ref{sec:revisited} overview results from running different bandit algorithms in isolation (i.e. without competition). Sections ~\ref{sec:competition} and ~\ref{sec:barriers} overview the results of the reputation choice variant. Section ~\ref{sec:non_greedy} presents results of varying the consumer choice rule in the reputation choice variant. Section ~\ref{sec:conclusion} concludes.
